---
title: Identification of common mistakes on performing a dumbbell lift using movement
  sensors
author: "Gonçalo Nogueira"
date: "August 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(dplyr)
library(randomForest)
```

## Summary

This study uses data collected by *Velloso et al* where four inertial measurement units
were placed on three different spots of the subject performing the exercise and on
the dumbell being raised. The raw data on the six subjects was preprocessed by the authors
by defining windows and calculating statistics on each of the signals provided by the
IMU's. The `classe` variable is the outcome and it can take different values based on
the technique of the subject performing the exercise.

Our approach started with some feature reduction, first by identifying covariates with near
zero variance and then by using a kind of wrapper method. We built a random forest model
on a subset of the training data using all variables. Then we calculated the importance
of each variable in the model and used the 10 most importante features in the final model.

The final model is a random forests model with an overall accuracy of approximately 98.4%. It is
very accurate at correctly identifying class A and class E cases, but not so accurate in other cases.

## Exploratory data analysis

```{r}
dataset <- read.csv("pml-training.csv")
final.testing <- read.csv("pml-testing.csv")

dim(dataset)
```
The dataset consists of 19622 observations with 160 variables. If one explores the
dataset, one will find some variables that were imported as factors, but are in fact numeric.
They were thus imported because some observations have the string `"#DIV/0"` as the value
of these variables. So we will replace this string with `NA` on these variables.

```{r, warning=FALSE}
factor.variables <- sapply(colnames(dataset), function(x) class(dataset[,x]) == "factor")
# There are some factor variables that we don't want to process in this way,
# so let's take them out of the list
factor.variables["user_name"] <- FALSE
factor.variables["classe"] <- FALSE
factor.variables["cvtd_timestamp"] <- FALSE
factor.variables["new_window"] <- FALSE
x <- apply(dataset[ ,factor.variables], 2, function(x) grep("#DIV/0!", x))
j <- 1
for (i in seq_along(factor.variables)) {
  if (factor.variables[i]) {
    dataset[x[[j]], i] <- NA
    f <- dataset[ , i]
    dataset[ , i] <- as.numeric(levels(f))[f]
    j <- j + 1
  }
}

proportion.of.nas.by.col <- apply(dataset, 2, function(x) sum(is.na(x)) / length(x))
length(proportion.of.nas.by.col[proportion.of.nas.by.col > .9])
```

As we can see above, 100 variables are `NA` on 90% or more observations. Obviously, these variables
are of no use in the training of a model, so we will filter these out.

```{r}
dataset <- dataset[, !(proportion.of.nas.by.col > .9)]
```

If we dig into the names of the variables, we can see that there are a few ones that do not make
sense in building this model. Specifically, the following variables probably won't add any
value to the model:

  * `X` - this is just an identifier of the observation; we already have the dataframe row
          number for that.
  * `raw_timestamp_part_1`, `raw_timestamp_part_2` and `cvtd_timestamp` - the time and date at which
          the exercise was done is of no interest to us.
  * `num_window` and `new_window` - this has to do with the way the researchers preprocessed the data
          and is of no interest to us.
  * `user_name` - since we're building a model that should apply to any subject, this variable can
  be discarded.
                  
So let's remove these variables from the dataset. Some predictor variables are integers but
this doesn't make much sense, so let's also convert all variables to numeric.

```{r}
dataset <- select(dataset,
                   -X,
                   -raw_timestamp_part_1,
                   -raw_timestamp_part_2,
                   -cvtd_timestamp,
                   -num_window,
                   -new_window,
                   -user_name)
dataset[ , -53] <- as.data.frame(sapply(dataset[,-53], as.numeric))
```

Now we're down to 52 predictors. We will build a random forests model on a subset of
10% of the observations of the dataset and then we will look into the importance of each of
the features in that model. Then we will use the 10 most important features to train the
final random forests model.

First let's partition our data into training and testing datasets. We will use 70% of the
data as the training portion and the rest as the testing portion.

```{r}
inTrain <- createDataPartition(dataset$classe, p = 0.7, list = FALSE)
training <- dataset[inTrain, ]
testing <- dataset[-inTrain, ]
```


```{r, cache = TRUE}
set.seed(1234)
for.feature.selection <- createDataPartition(training$classe, p = 0.1, list = FALSE)
feature.selection.subset <- training[for.feature.selection, ]
model <- randomForest(classe ~ ., feature.selection.subset)
variable.importance <- varImp(model)
variable.importance <- data.frame(variable = rownames(variable.importance),
                                  importance = variable.importance,
                                  stringsAsFactors = FALSE)
predictors.final.model <- head(arrange(variable.importance, desc(Overall)), n = 10)
predictors.final.model
```
Now that we have a reasonable subset of variables, we will subset the training dataset and build
the final random forests model.

```{r, cache=TRUE}
training <- select(training, predictors.final.model$variable, classe)
testing <- select(testing, predictors.final.model$variable, classe)
final.model <- train(classe ~ ., training,  method = "rf")
```

We will now evaluate our model by computing a confusion matrix and the overall accuracy of the model
using the testing partition of the original dataset.

```{r}
predictions <- predict(final.model, testing)
confusionMatrix(predictions, testing$classe)
```

So we have an overall accuracy that falls within the 95% confidence interval 98.05% to 98.71%. Looking
at the sensitivity statistics, we can see that the model is particularly accurate at predicting a
class A, D or E case (sensitivity is above or near 99%) but is a bit weaker at predicting other classes, especially class B which presents sensitivity of only 96.8%.

Let's now calculate the predictions for the final testing dataset of 20 observations that will
have to be inputted in the quiz.

```{r}
final.predictions <- predict(final.model, final.testing)
final.predictions
```